---
title: "Pre-processing with phyloseq"
author: Chloe J McGovern
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # send any figure output to this folder 
                      fig.path = "../figures/02_PreProcessing/") 
```


# Goals

Remove any potential contaminants and evaluate the accuracy of our sequencing run. Then we will write our final data objects. 

1. Load in data that we've generated: `asv_table` , `tax_table` , and metadata (which includes `metadata.csv` and `track_reads_reads`). We created these fines in `analysis/02_AssignASVs.Rmd`.
2. Remove ASVs that are chloroplasts.
3. Remove ASVs that are mitochondria.
~Endosymbiosis theory but still contamination~
4. Evaluate any ASVs from the negative controls. Then, we will remove the negative controls. 
5. Evaluate the mock community (postive control). Accuracy of sequencing. 
6. Check fro reverse compliments (can also be done in the filter and trim step).
7. Check the sequencing depth of samples. Remove obvious samples with too few reads. 

## Input

1. Metadata: `metadata.csv` and `trach_read_counts.RData`.
2. ASV table: `data/01_DADA2/ASV_table.csv`.
3. Taxonomy table: `data/01_DADA2.ASV_taxonomy.tsv`.

## Output

1. A pre-processed phyloseq object (S4 object): `raw_preprocessed_physeq.RData`.


# Set environment

## Load packages

```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")

# Load packages with pacman
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, install=FALSE)
```


## Load data

### 1. Metadata

-What to copy
   1. `/workdir/in-class-data/metadata.csv`
        -Where to copy? 
            `/workdir/cjm423/SalinityGradient_16S/data`
            
   2. `/workdir/in-class-data/track_read_counts.RData`
        -Where to copy?
            `/workdir/cjm423/SalinityGradient_16S/data/01_DADA2/`

```{r load-metadata}
metadata_df <- read_csv("data/metadata.csv") %>%
  # Fix column names
  dplyr::rename("sample_names" = "...1") %>%
  #Add sample names also as a column
  mutate(names = sample_names) 

#Inspect
head(metadata_df)
dim(metadata_df)

# include dada2 output

load("data/01_DADA2/track_read_counts.RData")

# Take a look
head(track_counts_df)
dim(track_counts_df)

#Check file names
head(track_counts_df$sample_names)

#Fix sample names in track_reads
track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names , "_"), `[`,1)

#Check
head(track_counts_df)

# What is different?
setdiff(track_counts_df$sample_names , metadata_df$sample_names)

metadata_final_df <- 
  metadata_df %>%
  left_join(., track_counts_df, by = "sample_names") %>%
  # we need our sample names in rownames! 
  column_to_rownames(var = "sample_names")

# Intuition check 
dim(metadata_final_df)

```

ASV table = ASV (row) x Sample name (c)
Taxonomy table = ASV (rows) x Taxonomy (col)

Need to do: Metadata = sample name (rows) x info (col)
*Seen in the chunk above as #Add sample names with `mutate()` function

## 2. ASV Table

```{r load-asv-table}

asv_df <- read.delim(file = "data/01_DADA2/ASV_table.tsv" , sep = "\t" ,
           #Add column names and row names
           header = TRUE , row.names = 1) %>%
  dplyr::select(-"CJ.V08.P")

#Inspect
asv_df[1:5, 1:5]

# Fix column names
## Remove the X
colnames(asv_df) <- sub("^X", "", colnames(asv_df))
## Replace the . with a - : "\\." since . is a special expression
colnames(asv_df) <- gsub(pattern = "\\." , replacement = "-" , colnames(asv_df))

#Check
head(colnames(asv_df))
asv_df[1:5,1:5]
```

## 3. Taxonomy Table

```{r load-tax-table}
tax_df <- read.delim("data/01_DADA2/ASV_taxonomy.tsv" , sep = "\t" ,
           header = TRUE , row.names = 1)

#Inspect
dim(tax_df)
dim(asv_df)

#Add check
stopifnot(row.names(asv_df) == row.names(tax_df))
```

# Handoff to phyloseq

```{r phyloseq-handoff}
raw_physeq <-
  phyloseq(otu_table(asv_df , taxa_are_rows = TRUE) , 
         sample_data(metadata_final_df) ,
         tax_table(as.matrix(tax_df)))

raw_physeq

# Save
save(raw_physeq , file = "data/03_PreProcessing/raw_physeq.RData")

```

# Clean the Data!

```{r rm-chloro}
noChloros_physeq <-
  raw_physeq %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order))

#How many ASVs were chloroplasts?
numChloros_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
```

There were `r numChloros_ASVs` ASVs that were chloroplasts.


```{r rm-mitos}
noChlorosMitos_physeq <-
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

#How many ASVs were chloroplasts?
numMitos_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
```

There were `r numMitos_ASVs` ASVs that were chloroplasts.


# Evaluate and remove the control samples

```{r neg-control}
#Make phyloseq object with only the negative controls

neg_control_samples <- c("WaterControl" , "022um-Control" , 
                         "3um-Control" , "DNA-Ext-Control")

#create phyloseq object with only negative controls
neg_control_physeq <-
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% neg_control_samples) %>%
  # rm ASVs with 0 counts using prune_taxa
  prune_taxa(taxa_sums(.)> 0, .)

# Take a look at ASV abundances in negative controls
neg_control_physeq %>%
  plot_bar(. , "ASV" , fill = "ASV")


#What is the abundance of the negative control ASVs in real samples?
#create a vector of ASVs in neg controls 
neg_control_ASVs_vec <-
  neg_control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()
  
#subset from larger dataset to create new phyloseq object
neg_controlASV_physeq <-
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% neg_control_ASVs_vec$ASV)
  

#Take a look
neg_controlASV_physeq

#View(metadata_final_df)

#plot abundances of neg control ASVs in sample vs control
neg_controlASV_physeq %>%
  plot_bar(., "ASV" , facet_grid = Sample_or_Control~.)

# Would only like to keep ASV_17 and ASV_50, everything else is trash
controlASVs_rm <- dplyr::setdiff(neg_control_ASVs_vec$ASV, c("ASV_0017", "ASV_0050"))
# double check 
length(controlASVs_toREMOVE)

#make new phyloseq object
noControlsChlorosMitos_physeq <- 
  noChlorosMitos_physeq %>%
  # remove the control ASVs (except 17, 50)
  subset_taxa(., !(ASV %in% controlASVs_rm)) %>%
  # Remove neg control samples 
  subset_samples(., !(names %in% neg_control_samples))

# Inspect no controls object 
noControlsChlorosMitos_physeq

```



