---
title: "04A_Phylogenetics_Tree_Construction"
author: "Chloe J McGovern"
date: "2025-03-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Goals

The goal is to make a phylogenetic tree!

1. Load in our pre-processed phyloseq object (raw_preprocessed_phyloseq.RData)
2. Create ASV Fasta file from the phyloseq object.
3. Align the 16S sequences from our fasta file with MAFFT
4. Create a maximum likelihood tree using FastTree2.

### Inputs
1. raw_preprocessed_phyloseq.RData

### Outputs
1. MAFFT_aligned_ASV.fasta
2. ASVs_unrooted.tree

# Set the Seed

```{r set-seed}
set.seed(238438)
```

# Load packages and Phyloseq Object
```{r load-packages}

# install phytools
#remotes::install_github("liamrevell/phytools")

#install ggtree
#devtools::install_github("GuangchuangYu/ggtree")

#Load into R session
pacman::p_load(phytools, ggtree, tidyverse, phyloseq, install=FALSE)

# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# Write ASV Fasta file
```{r create-preprocessed-fasta}
# pull out ASV sequences and ASV names
asv_seq_df <-
  raw_preprocessed_physeq@tax_table %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# add the ">" to make the fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

# create fasta object
asv_seq_fasta <-
  c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))

#write to a new fasta file
write(asv_seq_fasta,
      file = "data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Align the 16S sequences from fasta file with MAFFT
-`engine.opts = 'l'`: gives us our original terminal base environment

To test in the shell directly from Rmd 
mac: command + option + enter 
Windows: control + alt + enter 
```{r run-mafft, engine='bash', engine.opts='-l'}
# write bash code to run MAFFT alignment

# provide path to mafft
export PATH=/programs/mafft/bin:$PATH #ctrl+alt+enter for windows

# change directories to provide the fasta file we made above
cd data/04_Phylogenetic_Tree

pwd

#set seed for consistency and reproducibility
RANDOM=238428

# MAFFT automatically knows that it's a nucleotide alignment 
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta

# Change back to the project directory 
cd ../../
pwd

```

# FastTree2
```{r run-fastTree, engine='bash', engine.opts='-l'}
# provide our export path for fast tree
export PATH=/programs/FastTree-2.1.11:$PATH

# change directory into the alignment folder
cd data/04_Phylogenetic_Tree

# run fastTree to generate phylogenetic tree
# parameters:
  # -nt = indicates it is a nucleotide alignment 
  # -gtr = generalized time reversible substitution model
  # -fastest = speed up the model and reduce memory usage (more than >50,000 seqs)
  # -log = output a log file (under the hood)
  # input our alignment 
  #specify the output tree file
 
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

#change back to main directory
cd ../../
pwd
echo "The working directory is $PWD"

```



