---
title: "04B_Phylogenetic_Tree_Inspection_Rooted"
author: "Chloe J McGovern"
date: "2025-03-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Goals
We want to investigate and root our tree

1. Load the FastTree unrooted
2. Add the tree to phyloseq object
3. Visualize and inspect tree with ggtree
4. Prune ASVs if needed
5. Root our tree
6. Combine enwly rooted tree with phyloseq object
7. Save 2 phyloseq objects
  -unrooted tree phyloseq object
  -rooted tree phyloseq object
  
# Input
1. raw_preprocessed_phyloseq.RData
2. ASVs_unrooted.tree
3. MAFFT_aligned_ASVs.fasta

# Output
1. phytree_preprocessed_physeq.RData

```{r set up, include=FALSE}
knitr::opts_chunk$set(echlo = TRUE,
                      fig.align = "center",
                      #Always relevant to the document directory
                      fig.path = "../figures/04_Phylogenetic_Tree")
```

# Before you start

## Set seed
```{r set-seed}
# any number 
set.seed(238438)
```

## Load packages
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, tidytree,
               install=FALSE)
```

## Load data files
```{r load-data}
#load preprocessed phyloseq object
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# load in tree
unrooted_tree <-
  read.tree("data/04_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree

# view structure of unrooted tree
str(unrooted_tree)

```

## Merge phyloseq object 
```{r merge-physeq}
#intuition check
stopifnot(ntaxa(raw_preprocessed_physeq)==ntaxa(unrooted_tree)) #no news is good news

# merge tree with phyloseq object
unrooted_physeq <-
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)

unrooted_physeq
```

## Plot tree with ggtree
```{r plot-tree-unrooted}
#make a basic tree
kingdom_tree <-
  ggtree(unrooted_physeq) +
  #color tips by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) +
  # add a title 
  labs (title = "Unrooted Tree") +
  #move the legend to the bottom
  theme(legend.position = "bottom")

kingdom_tree

#view kingdom with labels
kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)

kingdom_node_tree
```

## Root tree from archea
NOTE: got to line 232 if you do not have archea in your dataset #that is me
```{r root-archea}
# first find the ASVs that are Archaea
archaea_node <- tax_table(raw_preprocessed_physeq) %>%
  as.data.frame() %>%
  filter(Kingdom == "Archaea") %>%
  pull(ASV)

# find the node that encompasses all Archaea
archaea_node <- findMRCA(unrooted_tree, tips = archaea_node, type = "node") # node = 3394

# View clade to check it out
#viewClade(kingdom_tree, archaea_node)
viewClade(kingdom_node_tree, 3394)

# check out tree
node3394 <-
  ggtree(unrooted_physeq) +
  geom_tippoint(mapping = aes(color = Kingdom),
                size = 1.5) +
  geom_text(aes(label = node), hjust = -0.7, vjust = -0.2, size = 2) +
  geom_highlight(node = 3394, fill = "dodgerblue")

node3394

# root tree using node 3394
archaea_root_tree <-
  ape::root(unrooted_tree, node = archaea_node, resolve.root = TRUE)

# check if rooted
is.rooted(archaea_root_tree)

# merge with physeq
archaea_root_physeq <- merge_phyloseq(raw_preprocessed_physeq, archaea_root_tree)

# plot rooted simple tree with no text
archaea_rooted_simple <- ggtree(archaea_root_physeq) +
  geom_tippoint(aes(color = Kingdom))  

archaea_rooted_simple

#plot with text
# Plot rooted tree with text
archaea_rooted_text <- ggtree(archaea_root_physeq) +
  geom_tippoint(aes(color = Kingdom)) + 
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.3, size = 1.5)

archaea_rooted_text
```

## Evaluate long branches
This part will get a little manual
```{r evaluate-archaea-rooted-long-branches}
#zoom in on node that looks potentially suspicious
viewClade(archaea_rooted_text + 
          labs(title = "Archaea Rooted Tree: Node 703"), 
          node = 703)

# Lets see its ancestors
ancestor(archaea_root_tree, 703)

# Or if there are two nodes that we can see we can find its MRCA (most recent common ancestor)
## create tibble of tree to identify offspring
tree_tib <- as_tibble(archaea_root_physeq@phy_tree)

# lets look at the really long branch with a bunch of NAs
MRCA(archaea_root_tree, .node1 = 703, .node2 = 702) #3316

viewClade(archaea_rooted_text, 3316)
```

