---
title: "Biodiversity across a salinity gradient"
author: "Chloe J McGovern"
date: "`r Sys.Date()"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/05_Biodiversity")
```

# Goals

Run an analysis on the within sample biodiversity of the samples in our project.

1. First calculate the Hill numbers od the samples using the iNEXT package.
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT.
3. Analyze the hill numbers as it relates to stations and salinity in (PSU).\
4. Answer the scientific question at hand.

## Inputs

1. `phytree_preprocessed_physeq`

## Outputs

1. Calculated biodiversity measures (Hill numbers) across every sample.
2. Figures to visualize these changes in biodiversity.
3. Statistical tests to actually show the canges and patters in biodiversity.

# Set up the environment

## Set the seed
```{r set-seed}
set.seed(238428)
```

## Load packages and colors

```{r load-packages-colors}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install=FALSE)

# load in colors for plotting
source("code/colors.R")

```

## Load in data

```{r}
load("data/04_Phylogenetic_Tree/phytree_preprocessed_physeqs.RData")

# take a look
archaeal_rooted_physeq

# intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq)) #max, min, medium

# make metadata df
metadata_df<- archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

glimpse(metadata_df)
```

# Calculate biodiversity with iNEXT

```{r iNEXT-biodiversity}
# prepare the input data 
  #species in rows, samples in columns

iNEXT_input_df <-
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# inspect
dim(iNEXT_input_df)
iNEXT_input_df[1:5,1:5]

# run iNEXT
  # remember to set the seed
  #ASVs in rows, samples in columns

#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0,1,2),
#                    datatype = "abundance") #takes a while to run

```

Since iNEXT takes a while ti run, we will load in the data that was calculated before

```{r load-iNEXT-data}
load("data/05_Biodiversity/iNEXT_data.RData")

# inspect
str(iNEXT_data)
typeof(iNEXT_data)
```

# Plot the rarefaction curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")

head(iNEXT_color_df)
```

Remember the output of iNEXT

q = 0 richness
q = 1 shannon
q = 2 simpson

```{r iNEXT-rarefaction}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$station_colors) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors) +
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq)),
                     guide = FALSE) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 

```

As we increase the q value, we are increasing the weight of the proportion. Dominant species are bring weighted more as `q` goes up

Inverse Simpson least sensitive to sequencing depth. But if we care about rare groups, richness and rarefaction curves

Finally, let’s draw some conclusions from the plot above (we discussed in class & I’ll summarize here):

The best gift of a the Hill Numbers: All three diversity measures are on the same x-axis: The Effective Number of Species or ASVs.

As q increases from 0 to 1 to 2, the abundance of the species is weighted more, so the rarefaction curve plateaus earlier. 

This is important because it indicates that richess is the most sensitive to sequencing depth! 

However, Shannon and even less sensitive is Simpson!

Most of the sample rarefaction curves appear to plateau but we need to look deeper.

There appears to be a sample with much lower richness than the rest of the dataset. Let’s keep our eyes on this sample.

# Manually plot the rarefaction

```{r manual-rarefaction}
str(iNEXT_data)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(station_colors = station_colors,
                          station = names(station_colors)),
            by = "station") %>%
  mutate(station = factor(station, levels = c("Copano West", "Copano East", 
                                              "Mesquite Bay", "Aransas Bay", 
                                              "Shipping Channel")))

# inspect
dim(iNEXT_manual_df)

# manually plot rarefaction
iNEXT_manual_df %>%
  #filter out rows that are calculated out by sequences so no extrapolated data
  dplyr::filter(Method == "Rarefaction") %>%
  # make plot!
  ggplot(aes(x = m, y = qD, color = station, group = names)) +
  geom_line() +
  #facet by station
  facet_grid(Order.q~station, scales = "free") +
  scale_color_manual(values = station_colors) +
  theme_bw() +
  labs(y = "Effective Number of ASVs", x = "Number of Sequences")+
  theme(legend.position = "bottom")

```

# Boxplot: Diversity by Station

## Categorical analysis

```{r make-observed-df}
obs_div_df <-
  iNEXT_manual_df %>%
  dplyr:: filter(Method == "Observed")

# check it out
glimpse(obs_div_df)

# plot box plots by station against diversity

obs_div_df %>%
  ggplot(aes (x = station, y = qD, fill = station, color = station)) +
  facet_wrap(.~Order.q, scales = "free_y") +
  geom_jitter() +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_color_manual(values = station_colors) +
  scale_fill_manual(values = station_colors) +
  labs(y = "Effective number of ASVs") +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1,
                                   vjust =1))

```

# Continuous Relationships - Salinity (PSU) vs Biodiversity

## Environmental Variables

```{r environ-check}
# create df of env variables
env_df <-
  metadata_df %>%
  dplyr::select(names, water_tempC:DO_mgL)

# pairwise interactions between env var
pairs(dplyr::select(env_df, -names), upper.panel = NULL)
```

## PSU

- practical salinity unit
- 1 PSU = 1 g salt per 1,000 g water = ppt

0 -> 30 (gradient)

fresh: 0 - 0.5
brackish: 0.5 - 30
saline: > 30

How does biodiversity change over the salinity gradient?

```{r plot-div-salinity-lm}
obs_div_df %>%
  ggplot(aes(x = salinity_psu, y = qD)) +
  facet_wrap(.~Order.q, scales = "free") +
  geom_point(aes(color = station))+
  scale_color_manual(values = station_colors) +
  labs(x = "Salinity (PSU)", y = "Effective number of ASVs") +
  stat_smooth(method = "lm") +
  theme_bw() +
  theme(legend.position = "bottom")

```
