---
title: "Between-Sample (Beta) Diversity of Microbes along a Saliity Gradient"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/06_CommunityAnalysis")
```


# Goals

1. Load in the phyloseq data with our rooted tree, colors, functions.
2. Evaluate sequencing depth and remove samples, if needed.
3. Normalize read counts between samples.
4. Calculate community **dis**similarities: Numbers will always mean 0 = the same and 1 = dissimilar or different.
  - Sorensen
  - Bray Curtis
  - Unifrac (weighted and unweighted)
5. Run some statistics: PERMANOVA and betadisper.
6. Plot some bar plots of specific microbial groups in the data.

## Inputs

1. Phylogenetic tree phyloseq object that we created in `../analysis/04B_PhylogeneticTree_Inspection.Rmd`.

## Outputs

1. Calculated beta diversity dissimilarity measures.
2. Figures: PCoA and NMDS plots.
3. Statistical tests: PERMANOVA results and post-hoc betadisper tests.

# Scientific Question

## How does microbial composition change across a salinity gradient? 

-*Null Hypothesis*: No change in microbial community with varying salinity.
-*Alternative Hypothesis*: Community composition will change with varying salt concentrations. 

The more exciting part is how the community changes
  - Dominant members? (**Bray Curtis**)
  - Rare groups? (Jaccard, a binary measure, **Sorensen**, shared species, **unweighted Unifrac** for rare and dominant)
  - Evolutionary related? (weighted Unifrac, branch length)

# Set up

## Set seed

```{r set-seed}

set.seed(238428)

```

## Load packages

```{r load-packages}

pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan,
               install = FALSE)

# Load colors
source("code/colors.R")

# Load functions
source("code/functions.R")

```

# 1. Load in the data

```{r load-data}
# Load phyloseq object

load("data/04_Phylogenetic_Tree/phytree_preprocessed_physeqs.RData")
archaeal_rooted_physeq

# Glance at sequence depth

min(sample_sums(archaeal_rooted_physeq)) # called minimum library size
max(sample_sums(archaeal_rooted_physeq))

# create metadata df

metadata_df <- 
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

```

# Normalize read depth

## 2. Explore raw read counts

```{r explore-read-counts}

# calculate read counts per sample

raw_TotalSeqs_df <- archaeal_rooted_physeq %>%
  # calc total seq/reads
  sample_sums() %>%
  data.frame()
  
head(raw_TotalSeqs_df)

# Rename the column
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs" # rerun head

# add a new column with the number of ASVs (raw, non-normalized number of ASVs)

raw_TotalSeqsASVs_df <-
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(archaeal_rooted_physeq) > 1))

# View(raw_TotalSeqsASVs_df)

# plot a histogram of seq depth

raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0, 7500)) + # over shoot max
  labs(title = "Raw Seq Depth") +
  theme_bw()

# plot seq depth vs number of ASVs

raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 7500)) +
  labs(title = "Seq Depth vs Num ASVs")

summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))

```

## 3. Scaling the read counts

```{r scale_reads}

min(sample_sums(archaeal_rooted_physeq)) # 1576

# scale the reads

scaled_physeq <- 
  archaeal_rooted_physeq %>%
  scale_reads(round = "matround")

scaled_physeq

# confirm seq read depth of scaled physeq

scaled_TotalSeqs_df <-
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"


# plot

scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0, 7500)) +
  theme_bw()

# Scaling always gives us a range!

min(scaled_TotalSeqs_df$TotalSeqs)
max(scaled_TotalSeqs_df$TotalSeqs)
range(scaled_TotalSeqs_df$TotalSeqs)

41/max(scaled_TotalSeqs_df$TotalSeqs)

```

# 4 - 6. Calculate and Visualize Community Dissimilarity, Statistical Significance, and Plots

Two different types of visualization

1. PCoA - Eigen-based analysis, linear, axis coefficients are meaningful because they represent the eigenvectors. 
2. NMDS - non-linear method where the axes are not meaningful, smush many axes of data into 2 axes

## Sorensen Dissimilarity Method

### Calculate dissimilarity

```{r soren-dis}

scaled_soren_dist <-
  phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

class(scaled_soren_dist)

# PERMANOVA : are the centroids different?

adonis2(scaled_soren_dist ~ station, data = metadata_df)

adonis2(scaled_soren_dist ~ station + date, data = metadata_df, by = "margin") # additive
adonis2(scaled_soren_dist ~ station * date, data = metadata_df, by = "margin") # interaction 

# post hoc betadisper

dispr_soren_station <- betadisper(scaled_soren_dist, metadata_df$station)
permutest(dispr_soren_station)

```

### Plot PCoA

```{r soren-pcoa}

# first we will calculate the pcoa with sorensen

scaled_soren_pcoa<-
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE) # false would be bray curtis,  true is sorensen

str(scaled_soren_pcoa)

# plot it

plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_pcoa,
                color = "station",
                title = "Sorenesen PCoA") +
  scale_color_manual(values = station_colors) +
  geom_point(size = 5, alpha = 0.5, aes(color = station)) +
  theme_bw() +
  theme(legend.position = "bottom")

```

### Plot NMDS

```{r soren-nmds}

scaled_soren_nmds<-
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = TRUE) # false would be bray curtis,  true is sorensen

plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_nmds,
                color = "station",
                title = "Sorenesen NMSDS") +
  scale_color_manual(values = station_colors) +
  geom_point(size = 5, alpha = 0.5, aes(color = station)) +
  theme_bw() +
  theme(legend.position = "bottom") 

# points that are close have more similar community structure from binary presence / absence

```




