---
title: "Microbial composition analysis along a saliity gradient"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/07_Composition")
```


# Goals

1. Load in scaled phyloseq object
2. Calculate relative abundance of taxonomic groups at various levels: Phylum, Family, Genus, ASV
3. Plot them and narrow in on specific taxonomic groups of interest

## Inputs

1. scaled phyloseq object

## Outputs

1. Beautiful visualizations of microbial taxa and how they vary across station and salinity
2. Run some stats


# Set the environment

## Load packages

```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork,
               install = FALSE)
```

## Load in data

```{r load-data-colors}
load("data/06_Ordination/scaled_physeq.RData")

# look 
scaled_physeq

# colors
source("code/colors.R")

#intuition check
min(sample_sums(scaled_physeq))
range(sample_sums(scaled_physeq)) #yay!
```

# Taxonomic Analysis 

## Phylum

Calculate the relative abundance of the phyla across all the samples.

NOTE: all the data must be normalized for comparison--here 1942.

```{r calc-phylum-df}

phylum_df <-
  scaled_physeq %>%
  # agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # relative abundance
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # create df from phyloseq object
  psmelt() %>%
  # filter out phyla that are less than 1%
  dplyr::filter(Abundance > 0.01) %>%
  # fix the order of date
  mutate(date = fct_relevel(date, c("6/2/21", "6/15/21", "10/5/21")),
         station = fct_relevel(station, c("Copano West", "Copano East",
                                          "Mesquite Bay", "Aransas Bay",
                                          "Shipping Channel")))
  

```

yay now we have a DF we can plot with! Time for Mar's favorite: a stacked bar plot

```{r stacked-bar-phylum}
phylum_df %>%
  #individual sample comparison of surface waters, whole fraction
  # this will allow us to look at individual sample
  dplyr::filter(depth == 0.0) %>%
  dplyr::filter(fraction == "Whole") %>%
  ggplot(aes(x = station, y = Abundance, fill = Phylum)) +
  facet_grid(Phylum~date) + 
  # add stacked bar
  geom_bar(stat = "identity" , color = "black") +
  # change colors to be selected colors
  scale_fill_manual(values = phylum_colors) 
  
```

# Actinomycetota

```{r phylum-actinos}



```

## Family

```{r}
phylum_df %>%
  dplyr::filter(Phylum == "Actinomycetota") %>%
  ggplot(aes(x = station, y = Abundance, color = station, fill = station)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter() +
  theme_bw() +
  scale_color_manual(values = station_colors) +
  scale_fill_manual(values = station_colors) +
  labs(title = "Actinomycetota Relative Abundance")
  
```

## Genus

```{r}

```

## ASV


